Discrete Quantum Theories and Computing

### 2018-12-25 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" Updating...

%% LyX 2.3.1-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,twoside]{iuphd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fontspec}
\setmainfont[Mapping=tex-text]{Times New Roman}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{rotfloat}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{graphicx}
\usepackage[all]{xy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\providecommand{\LyX}{L\kern-.1667em\lower.25em\hbox{Y}\kern-.125emX\@}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how to, classical or quantum, attack these two tasks, they
require two different types of quantities. On one hand, a question
and its answer of RSA are both discrete integers. On the other hand,
molecules are described by quantities including the domain of their
position wave functions or the probability amplitudes of a wave function.
Even if these quantities look like continuous, we want to argue that
either these quantities might later be discovered to be discrete,
or the exact value of these continuous quantities may never be pinned
down. To better understand this issue without struggling with the
quantum theory, it is instructive to review how classical quantities
are characterized and computed.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick are made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
actually be an integer multiple of the elementary charge. A prefect
square cannot be draw physically. Even we don't draw it, but merely
ask whether it exists in the space. This question is still beyond
our ability to answer because whether the space itself is ultimately
discrete or continuous is still an unsolved question. Even if a prefect
square exists physically, we may not able to decide whether the ``square''
we discussed is really prefect or almost prefect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation is closed
to zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by an analog computer. Since two branches of computations
may not have any relations, the later computational path may be completely
different like the butterfly effect, and this kind of difference cannot
be compensated by error analysis techniques easily. Therefore, the
computational power of analog computers might be far weaker than the
prediction based on this kind of theoretical models.

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{ccYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but after we have a better technology to manipulate the minimum units, we can predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum models &  & Quantum Theories and Computing over Finite Fields & Quantum Interval-Valued Probability Measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and the
two possibilities dealing with real-valued models inspires us two
types of quantum models. Similar to classical analog computers, the
quantum circuit model, the most widely used model of quantum computing,
manipulates quantities which are not discrete apparently: the probability
amplitudes of quantum states. To address the similar inconsistency,
we first review the ``conventional quantum theory''\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.} in Chapter~\ref{chap:Conventional-Quantum-Theory}, where the probability
amplitudes are in the field of complex numbers. If the probability
amplitudes are assumed to be ultimately discrete, we replace the field
of complex numbers by discrete finite fields and consider the quantum
theories and computing over finite fields. Another possibilities could
be that the probability amplitudes are never quantized no matter how
hard we try, and we could never access their exactly values. Notice
that the probability amplitudes can never be accessed directly in
an experiment, the accessible properties of a quantum system is its
measurement results and the probability of an event. Hence, to incorporate
the finite precision measurement into the quantum theory, we consider
a framework called quantum interval-valued probability measures (QIVPM).

@@

\begin{figure}[tb]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The two-sphere~$\Sphere{2}$ represented by \Eq{A7.eq}, which
is the irreducible space of one-qubit states, along with a representative
set of points on the sphere. Each single point on the sphere in (a)
corresponding to a circle in (b), and a whole family of circles (the
paths of $\rme^{\rmi\theta}$) on the three-sphere~$\Sphere{3}$
represents the Hopf fibration, Eq.~(\ref{eq:HopfFibration}). Although
$\Sphere{3}$ cannot be directly embedded in $\R^{3}$, three-sphere~$\Sphere{3}$
can be regarded as attaching two three-dimensional ball on two sides
of two-sphere~$\Sphere{2}$. In this way, each circle in $\Sphere{3}$
can be represented as a circle in the three-dimensional ball as shown
in (b). Moreover, points in (a) are color coded corresponding to circles
in (b), e.g., one pole contains the red elliptical circle that would
become an infinite-radius circle by a slightly different way to represent
$\Sphere{3}$ in $\R^{3}$, and the opposite pole corresponds to the
large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}
The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\left(\theta+\phi\right)\cos\psi,\,\sin\left(\theta+\phi\right)\cos\psi,\,\cos\left(\theta-\phi\right)\sin\psi,\,\sin\left(\theta-\phi\right)\sin\psi\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\frac{\pi}{2}$, with $0\leq\theta+\phi<2\pi$ and $0\leq\theta-\phi<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\left(2\phi\right)\sin\left(2\psi\right),\,\sin\left(2\phi\right)\sin\left(2\psi\right),\,\cos\left(2\psi\right)\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to $\ket{\psi_{1}}=\rme^{2\rmi\phi}\cos\psi\ket{0}+\sin\psi\ket{1}$,
and an irreducible state can be represented as a point on a sphere
called the Bloch sphere, as shown in Fig.~\ref{blochSphere.fig}(a).

\begin{figure}[tb]
\begin{centering}
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.47\columnwidth]{blochSphere}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphereArc}
\par\end{centering}
}\hfill{}
\par\end{centering}
\caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}
Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Hat{a}$ and $\Hat{b}$ on $\Sphere{2}$,
the shortest rotation carrying $\Hat{a}$ to $\Hat{b}$ is the SLERP
(spherical linear interpolation) \cite{Shoemake:1985:ARQ:325334.325242,wiki:Slerp}
\begin{equation}
S\left(\Hat{a},\Hat{b},t\right)=\Hat{a}\,\frac{\sin((1-t)\omega)}{\sin\omega}\,+\,\Hat{b}\,\frac{\sin(t\omega)}{\sin\omega}\,,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\omega$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

\begin{figure}[t]
\[
\xyR{.9em}\xyC{.6em}\entrymodifiers={@*=<0em>}\xymatrix{\lstick{y=\ket{0}} & \qw & \multigate{3}{\uf} & \gate{~S^{\dagger}~} & \ctrl{3} & \gate{~S^{\dagger}~} & \multimeasureD{3}{\text{measure}}\\
\lstick{x_{1}=\ket{0}} & \multigate{2}{\otimes S} & \ghost{\uf} & \multigate{2}{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{\ldots} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{x_{n}=\ket{0}} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}
}
\]
\caption{\label{fig:alg}Circuit for black box \usat\ in modal quantum theory
over the field $\ff{2}$. For further notation see text.}
\end{figure}
We can prove the unreasonable power of the arbitrary-function \usat\ starting
with a classical function $f:\boolt^{n}\rightarrow\boolt$ that takes
$n$ bits and returns at most one \btrue\ result. To build a quantum
algorithm, $f$ is first represented as the Deutsch quantum black
box~$\uf$ with \cite{Deutsch1985,544199} 
\begin{equation}
\uf\ket{y}\ket{\overline{x}}=\ket{y\oplus f\left(\overline{x}\right)}\ket{\overline{x}}=\begin{cases}
\ket{y}\ket{\overline{x}} & \textrm{if }f\left(\overline{x}\right)=\bfalse\,;\\
\ket{\mathsf{not}\left(y\right)}\ket{\overline{x}} & \textrm{if }f\left(\overline{x}\right)=\btrue\,,
\end{cases}\label{eq:DeutschBox}
\end{equation}
where $\overline{x}$ denotes a sequence $x_{1},x_{2},\ldots,x_{n}$
of $n$ bits, $\oplus$ is exclusive disjunction, and $0$ and $1$
are identified as $\bfalse$ and $\btrue$, respectively. Then, we
can give an algorithm (see Fig.~\ref{fig:alg}) taking as input such
a classical function that decides, deterministically and in a constant
number of black box evaluations, whether $f$ is satisfiable or not:

@@

\centering{}%
\begin{tabular}{ccccccccc}
\toprule 
\addlinespace
$j$ & $0$ & $1$ & $2$ & $3$ & $4$ & $5$ & $6$ & $7$\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
$\left(1+\rmi\right)^{j}$ & $1$ & $1+\rmi$ & $-\rmi$ & $1-\rmi$ & $-1$ & $-1-\rmi$ & $\rmi$ & $-1+\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(1-\rmi\right)^{j}$ & $1$ & $1-\rmi$ & $\rmi$ & $1+\rmi$ & $-1$ & $-1+\rmi$ & $-\rmi$ & $-1-\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(-1+\rmi\right)^{j}$ & $1$ & $-1+\rmi$ & $\rmi$ & $-1-\rmi$ & $-1$ & $1-\rmi$ & $-\rmi$ & $1+\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(-1-\rmi\right)^{j}$ & $1$ & $-1-\rmi$ & $-\rmi$ & $-1+\rmi$ & $-1$ & $1+\rmi$ & $\rmi$ & $1-\rmi$\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}
In Table~\ref{tab:Generators-in-F9}, one can notice that $\left(a+b\rmi\right)^{3}=a-b\rmi$.
In general, the $p$-th power $\left(a+b\rmi\right)^{p}=a-b\rmi$
is called the \emph{Frobenius automorphism} acts like complex conjugation
$\left(a+b\rmi\right)^{*}=a-b\rmi$ \cite{Artin1991,grove2002classical,DummitFoote2004}.
Then, we define the \emph{field norm} $\fnorm{\cdot}:\Fpp\rightarrow\Fp$
as an element $a+\rmi b$ multiplying its complex conjugation $\left(a+b\rmi\right)^{*}$~\cite{wiki:FieldNorm},
\begin{equation}
\fnorm{a+\rmi b}=\left(a+b\rmi\right)\left(a+b\rmi\right)^{*}=\left(a+b\rmi\right)^{p+1}=a^{2}+b^{2}\,,\label{fieldnorm.eq}
\end{equation}
where the square root in the usual definition of norm is avoided because,
unlike the continuous case, the square root does not always exist,
and the field norm of an element $\fnorm{\cdot}$ should be the direct
counterpart of the norm-squared $\left|\cdot\right|^{2}$ in the conventional
quantum theory. For example, the field norm of every generator~$g$
in Table~\ref{tab:Generators-in-F9} is the same number $\fnorm{g}=g^{3+1}=-1$.
In fact, these four generators are the only elements in $\ff{3^{2}}$
whose field norm is $-1\in\ff{3}$. Generally, given any $c\in\Fp$,
let $\fnorm*{\left\{ c\right\} }$ denote the set of element whose
field norm is $c$, i.e., $\fnorm*{\left\{ c\right\} }=\set{\alpha\in\Fpp}{\fnorm{\alpha}=c}$.
The set $\fnorm*{\left\{ c\right\} }$ is the discrete analog of phase-equivalence
under the modulus-preserving transformation $z\rightarrow\rme^{\rmi\phi}z$,
and the number of its elements is characterized by the following proposition.

@@

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{prop_letter_ajhs_referee_response_nb} 
\par\end{centering}
\caption{The region to the left of the vertical line at $\delta=\frac{1}{3}$
is where we assume small measurement degradation; in that region our
extension of the KS theorem definitely demonstrates contextuality
(\textsf{C}). In the region to the right, the degradation of the data
is large and our extension of the KS theorem no longer refutes other
explanations for the experimental data.}
\label{fignoname} 
\end{figure}
As is the case for conventional, infinitely-precise, quantum probability
measures, the theorem is only applicable to dimensions $D\geq3$.
Indeed when the Hilbert space has dimension 2, it is straightforward
to construct a 0-deterministic QIVPM as follows. Consider a non-contextual
hidden variable model for $D=2$ (e.g., as proposed by Bell or Kochen-Specker~\cite{BELL_1966,kochenspecker1967}).
Such a two-dimensional model assigns definite values to all observables
at all times, and hence assigns a \emph{determinate} probability (0
or 1) to each event. This probability measure directly induces a 0-deterministic
QIVPM by changing 0 to $\imposs$ and 1 to $\necess$. It follows
that every 0-deterministic QIVPM is $\delta$-deterministic.

>>>>

%% LyX 2.3.2-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,twoside]{iuphd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fontspec}
\setmainfont[Mapping=tex-text]{Times New Roman}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{verbatim}
\usepackage{rotfloat}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{pdfpages}
\usepackage{graphicx}
\usepackage[all]{xy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\providecommand{\LyX}{L\kern-.1667em\lower.25em\hbox{Y}\kern-.125emX\@}
\let\@@LyX\LyX
\def\LyX{\@ensure@LTR{\@@LyX}}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how to, classical or quantum, attack these two tasks, they
require two different types of quantities. On one hand, a question
and its answer of RSA are both discrete integers. On the other hand,
molecules are described by quantities including the domain of their
position wave functions and the probability amplitudes of a wave function.
Even if these quantities look like continuous, we want to argue that
either these quantities might later be discovered to be discrete,
or the exact value of these continuous quantities may never be pinned
down. To better understand this issue without struggling with the
quantum theory, it is instructive to review how classical quantities
are characterized and computed.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick are made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
actually be an integer multiple of the elementary charge. A prefect
square cannot be draw physically. Even we don't draw it, but merely
ask whether it exists in the space. This question is still beyond
our ability to answer because whether the space itself is ultimately
discrete or continuous is still an unsolved question. Even if a prefect
square exists physically, we may not able to decide whether the ``square''
we discussed is really prefect or almost prefect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation is closed
to zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the following computational
path may be completely different like the butterfly effect, and this
kind of difference cannot be compensated by error analysis techniques
easily. Therefore, the computational power of analog computers might
be far weaker than the prediction based on this kind of theoretical
models. \todo{Maybe extended to consider probabilistic model? }

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{ccYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but after we have a better technology to manipulate the minimum units, we can predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum models &  & Quantum Theories and Computing over Finite Fields & Quantum Interval-Valued Probability Measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities dealing with real-valued models inspires us two types
of quantum models. Similar to classical analog computers, the quantum
circuit model, the most widely used model of quantum computing, manipulates
quantities which are not discrete apparently: the probability amplitudes
of quantum states. To address the similar inconsistency, we first
review the ``conventional quantum theory''\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.} in Chapter~\ref{chap:Conventional-Quantum-Theory}, where the probability
amplitudes are in the field of complex numbers. If the probability
amplitudes are assumed to be ultimately discrete, we replace the field
of complex numbers by discrete finite fields and consider the quantum
theories and computing over finite fields. Another possibilities could
be that the probability amplitudes are never quantized no matter how
hard we try, and we could never access their exactly values. Notice
that the probability amplitudes can never be accessed directly in
an experiment, the accessible properties of a quantum system is its
measurement results and the probability of an event. Hence, to incorporate
the finite precision measurement into the quantum theory, we consider
a framework called quantum interval-valued probability measures (QIVPM).

@@

\begin{figure}[tb]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The two-sphere~$\Sphere{2}$ represented by \Eq{A7.eq}, which
is the irreducible space of one-qubit states, along with a representative
set of points on the sphere. Each single point on the sphere in (a)
corresponding to a circle in (b), and a whole family of circles (the
paths of $\rme^{\rmi\theta}$) on the three-sphere~$\Sphere{3}$
represents the Hopf fibration, Eq.~(\ref{eq:HopfFibration}). Although
$\Sphere{3}$ cannot be directly embedded in $\R^{3}$, three-sphere~$\Sphere{3}$
can be regarded as attaching two three-dimensional ball on two sides
of two-sphere~$\Sphere{2}$. In this way, each circle in $\Sphere{3}$
can be represented as a circle in the three-dimensional ball as shown
in (b). Moreover, points in (a) are color coded corresponding to circles
in (b), e.g., one pole contains the red elliptical circle that would
become an infinite-radius circle by a slightly different way to represent
$\Sphere{3}$ in $\R^{3}$, and the opposite pole corresponds to the
large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\left(\theta+\phi\right)\cos\psi,\,\sin\left(\theta+\phi\right)\cos\psi,\,\cos\left(\theta-\phi\right)\sin\psi,\,\sin\left(\theta-\phi\right)\sin\psi\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\frac{\pi}{2}$, with $0\leq\theta+\phi<2\pi$ and $0\leq\theta-\phi<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\left(2\phi\right)\sin\left(2\psi\right),\,\sin\left(2\phi\right)\sin\left(2\psi\right),\,\cos\left(2\psi\right)\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to $\ket{\psi_{1}}=\rme^{2\rmi\phi}\cos\psi\ket{0}+\sin\psi\ket{1}$,
and an irreducible state can be represented as a point on a sphere
called the Bloch sphere, as shown in Fig.~\ref{blochSphere.fig}(a).

\begin{figure}[tb]
\begin{centering}
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.47\columnwidth]{blochSphere}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphereArc}
\par\end{centering}
}\hfill{}
\par\end{centering}
\caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Hat{a}$ and $\Hat{b}$ on $\Sphere{2}$,
the shortest rotation carrying $\Hat{a}$ to $\Hat{b}$ is the SLERP
(spherical linear interpolation) \cite{Shoemake:1985:ARQ:325334.325242,wiki:Slerp}
\begin{equation}
S\left(\Hat{a},\Hat{b},t\right)=\Hat{a}\,\frac{\sin((1-t)\omega)}{\sin\omega}\,+\,\Hat{b}\,\frac{\sin(t\omega)}{\sin\omega}\,,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\omega$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

\begin{figure}[t]
\[
\xyR{.9em}\xyC{.6em}\entrymodifiers={@*=<0em>}\xymatrix{\lstick{y=\ket{0}} & \qw & \multigate{3}{\uf} & \gate{~S^{\dagger}~} & \ctrl{3} & \gate{~S^{\dagger}~} & \multimeasureD{3}{\text{measure}}\\
\lstick{x_{1}=\ket{0}} & \multigate{2}{\otimes S} & \ghost{\uf} & \multigate{2}{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{\ldots} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}\\
\lstick{x_{n}=\ket{0}} & \ghost{\otimes S} & \ghost{\uf} & \ghost{\otimes S} & \targ & \qw & \ghost{\text{measure}}
}
\]
\caption{\label{fig:alg}Circuit for black box \usat\ in modal quantum theory
over the field $\ff{2}$. For further notation see text.}
\end{figure}

We can prove the unreasonable power of the arbitrary-function \usat\ starting
with a classical function $f:\boolt^{n}\rightarrow\boolt$ that takes
$n$ bits and returns at most one \btrue\ result. To build a quantum
algorithm, $f$ is first represented as the Deutsch quantum black
box~$\uf$ with \cite{Deutsch1985,544199} 
\begin{equation}
\uf\ket{y}\ket{\overline{x}}=\ket{y\oplus f\left(\overline{x}\right)}\ket{\overline{x}}=\begin{cases}
\ket{y}\ket{\overline{x}} & \textrm{if }f\left(\overline{x}\right)=\bfalse\,;\\
\ket{\mathsf{not}\left(y\right)}\ket{\overline{x}} & \textrm{if }f\left(\overline{x}\right)=\btrue\,,
\end{cases}\label{eq:DeutschBox}
\end{equation}
where $\overline{x}$ denotes a sequence $x_{1},x_{2},\ldots,x_{n}$
of $n$ bits, $\oplus$ is exclusive disjunction, and $0$ and $1$
are identified as $\bfalse$ and $\btrue$, respectively. Then, we
can give an algorithm (see Fig.~\ref{fig:alg}) taking as input such
a classical function that decides, deterministically and in a constant
number of black box evaluations, whether $f$ is satisfiable or not:

@@

\centering{}%
\begin{tabular}{ccccccccc}
\toprule 
\addlinespace
$j$ & $0$ & $1$ & $2$ & $3$ & $4$ & $5$ & $6$ & $7$\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
$\left(1+\rmi\right)^{j}$ & $1$ & $1+\rmi$ & $-\rmi$ & $1-\rmi$ & $-1$ & $-1-\rmi$ & $\rmi$ & $-1+\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(1-\rmi\right)^{j}$ & $1$ & $1-\rmi$ & $\rmi$ & $1+\rmi$ & $-1$ & $-1+\rmi$ & $-\rmi$ & $-1-\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(-1+\rmi\right)^{j}$ & $1$ & $-1+\rmi$ & $\rmi$ & $-1-\rmi$ & $-1$ & $1-\rmi$ & $-\rmi$ & $1+\rmi$\tabularnewline\addlinespace
\addlinespace
\addlinespace
$\left(-1-\rmi\right)^{j}$ & $1$ & $-1-\rmi$ & $-\rmi$ & $-1+\rmi$ & $-1$ & $1+\rmi$ & $\rmi$ & $1-\rmi$\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}

In Table~\ref{tab:Generators-in-F9}, one can notice that $\left(a+b\rmi\right)^{3}=a-b\rmi$.
In general, the $p$-th power $\left(a+b\rmi\right)^{p}=a-b\rmi$
is called the \emph{Frobenius automorphism} acts like complex conjugation
$\left(a+b\rmi\right)^{*}=a-b\rmi$ \cite{Artin1991,grove2002classical,DummitFoote2004}.
Then, we define the \emph{field norm} $\fnorm{\cdot}:\Fpp\rightarrow\Fp$
as an element $a+\rmi b$ multiplying its complex conjugation $\left(a+b\rmi\right)^{*}$~\cite{wiki:FieldNorm},
\begin{equation}
\fnorm{a+\rmi b}=\left(a+b\rmi\right)\left(a+b\rmi\right)^{*}=\left(a+b\rmi\right)^{p+1}=a^{2}+b^{2}\,,\label{fieldnorm.eq}
\end{equation}
where the square root in the usual definition of norm is avoided because,
unlike the continuous case, the square root does not always exist,
and the field norm of an element $\fnorm{\cdot}$ should be the direct
counterpart of the norm-squared $\left|\cdot\right|^{2}$ in the conventional
quantum theory. For example, the field norm of every generator~$g$
in Table~\ref{tab:Generators-in-F9} is the same number $\fnorm{g}=g^{3+1}=-1$.
In fact, these four generators are the only elements in $\ff{3^{2}}$
whose field norm is $-1\in\ff{3}$. Generally, given any $c\in\Fp$,
let $\fnorm*{\left\{ c\right\} }$ denote the set of element whose
field norm is $c$, i.e., $\fnorm*{\left\{ c\right\} }=\set{\alpha\in\Fpp}{\fnorm{\alpha}=c}$.
The set $\fnorm*{\left\{ c\right\} }$ is the discrete analog of phase-equivalence
under the modulus-preserving transformation $z\rightarrow\rme^{\rmi\phi}z$,
and the number of its elements is characterized by the following proposition.

@@

\begin{figure}
\begin{centering}
\includegraphics[scale=0.5]{prop_letter_ajhs_referee_response_nb} 
\par\end{centering}
\caption{The region to the left of the vertical line at $\delta=\frac{1}{3}$
is where we assume small measurement degradation; in that region our
extension of the KS theorem definitely demonstrates contextuality
(\textsf{C}). In the region to the right, the degradation of the data
is large and our extension of the KS theorem no longer refutes other
explanations for the experimental data.}
\label{fignoname} 
\end{figure}

As is the case for conventional, infinitely-precise, quantum probability
measures, the theorem is only applicable to dimensions $D\geq3$.
Indeed when the Hilbert space has dimension 2, it is straightforward
to construct a 0-deterministic QIVPM as follows. Consider a non-contextual
hidden variable model for $D=2$ (e.g., as proposed by Bell or Kochen-Specker~\cite{BELL_1966,kochenspecker1967}).
Such a two-dimensional model assigns definite values to all observables
at all times, and hence assigns a \emph{determinate} probability (0
or 1) to each event. This probability measure directly induces a 0-deterministic
QIVPM by changing 0 to $\imposs$ and 1 to $\necess$. It follows
that every 0-deterministic QIVPM is $\delta$-deterministic.
